---
title: "Ch 2 HCC Model"
author: "Tara Abrina & Oka Espenilla"
format: html
editor: visual
---

## Chapter 2 Hard Coral Cover Model

Version 3: refactored to use linear algebra for optimization and added 7 scenarios

## Required Packages

```{r}
if (!require(dplyr)) install.packages("dplyr"); library(dplyr)
if (!require(ggplot2)) install.packages("ggplot2"); library(ggplot2)
if (!require(tidyr)) install.packages("tidyr"); library(tidyr)
if (!require(data.table)) install.packages("data.table"); library(data.table)
if (!require(openxlsx)) install.packages("openxlsx"); library(openxlsx)
if (!require(igraph)) install.packages("igraph"); library(igraph)
if (!require(parallel)) install.packages("parallel"); library(parallel)
if (!require(foreach)) install.packages("foreach"); library(foreach)
if (!require(doParallel)) install.packages("doParallel"); library(doParallel)
```

## Data Wrangling

### Load data

Note: Make sure data files are in a folder called "data" in the project folder.

This code should run as long as ReefIDs, the data length and order in rows and columns are not changed. Underlying data can be changed to fit actual starting conditions in the future.

```{r}
data_path <- "data/"
output_path <- "output/"
nodes <- read.csv(paste(data_path,"centroid.csv",sep="")) #2404 ReefIDs with HCC, reef area, type, gravity penalty, mortality rate
edges <- read.csv(paste(data_path,"sourcesink.csv",sep="")) #full ReefID source-sink pairs with settlement probabilities
reeftype <-read.csv(paste(data_path,"type.csv",sep="")) #HACC & intrinsic growth per year by reef type
mmgrow <-read.csv(paste(data_path,"composition.csv",sep="")) #HACC & intrinsic growth per year by reef type in mm/year
preselect <- read.csv(paste(output_path,"preselect-ids.csv",sep="")) #top reefids based on net hcc
```

### Data Cleaning and International Reef Handling

```{r}
setnames(nodes, 
         old=c("ReefID","ave_hcc", "ReefArea.km.sq.", "type_assumption"), 
         new=c("reefid","hcc", "area_kmsq", "type"), skip_absent = TRUE)

setnames(edges, old=c("Probability.of.Settlement"), new=c("p_settle"), skip_absent = TRUE)

nodes$reefid <- as.character(nodes$reefid)
edges <- data.frame(
  source   = as.character(edges$SourceReef),
  sink     = as.character(edges$SinkReef),
  p_settle = as.numeric(edges$p_settle)
)

# Combine and Join Parameters
mmgrow$mmyr <- as.numeric (mmgrow$mmyr) #growth in mm as numeric
mmgrow_agg = mmgrow %>% group_by(type) %>% summarise(mmyr = sum(mmyr * percont), bleachmort = sum(bleachmort * percont), typhmort = sum(typhmort * percont), .groups = 'drop')
reef_params <- mmgrow_agg %>% left_join(reeftype, by = "type")

all_edge_reefs <- unique(c(edges$source, edges$sink)) #lists all nodes affecting PH nodes

international_nodes <- data.frame(reefid = setdiff(all_edge_reefs, nodes$reefid), area_kmsq = 1, type = "I", hcc = 1, penalty = 0, typhprob = 0, bleachprob = 0, is_philippines = FALSE)

nodes$is_philippines <- TRUE

nodes_final <- bind_rows(nodes, international_nodes) %>% left_join(reef_params, by = "type") %>%
  mutate(mmyr = ifelse(is.na(mmyr), 18.3, mmyr), typhmort = ifelse(is.na(typhmort), 0.02, typhmort), bleachmort = ifelse(is.na(bleachmort), 0.02, bleachmort),
         K = ifelse(is.na(hacc), 57.2, hacc), growthrate = ifelse(is.na(growthrate), 4.9, growthrate)/100) %>% distinct(reefid, .keep_all = TRUE)
```

### ! Target ReefIDs for Group Intervention

Please list ReefIDs for group intervention (default:active candidates)

```{r}
#preselected_ids <- c("2566","2539")

preselected_ids <- preselect |>
  filter(act. > 0) |>
  pull(reefid)
target_indices <- which(nodes_final$reefid %in% preselected_ids)
```

### Build Network and Extract Matrices

```{r}
coral_graph <- graph_from_data_frame(d = edges, vertices = nodes_final, directed = TRUE)
adj_mat <- as_adjacency_matrix(coral_graph, attr = "p_settle", sparse = TRUE)
P_T <- Matrix::t(adj_mat)

REEF      <- V(coral_graph)$name
HCC_INIT  <- V(coral_graph)$hcc
AREA      <- V(coral_graph)$area_kmsq
PENALTY   <- V(coral_graph)$penalty
K_VALS    <- V(coral_graph)$K
R_GROW    <- V(coral_graph)$growthrate
IS_PH     <- V(coral_graph)$is_philippines
```

Expected Annual Mortality (deterministic approach for speed)

```{r}
EXP_MORT  <- pmin(0.95, (V(coral_graph)$typhprob * V(coral_graph)$typhmort) +
                    (V(coral_graph)$bleachprob * V(coral_graph)$bleachmort))
```

Constants

```{r}
LARVAE_PER_HCC <- 1000000
MATURITY_RATE  <- 0.09
COLONY_TO_KMSQ <- 0.0000000002
STEPS <- 25
RESTORATION_LARVAE <- 1.4e9
```

## Function & For-Loop

```{r}

run_core_logic_ts <- function(hcc_start, steps, p_vec, r_indices=NULL, r_larvae=0) {
  ts_matrix <- matrix(0, nrow = length(hcc_start), ncol = steps)
  current_hcc <- hcc_start
  
  for (t in 1:steps) {
    larvae <- current_hcc * LARVAE_PER_HCC
    if (t == 1 && !is.null(r_indices)) {
      larvae[r_indices] <- larvae[r_indices] + r_larvae
    }
    
    settled <- as.vector(P_T %*% larvae)
    recruit <- ((settled * MATURITY_RATE * COLONY_TO_KMSQ) / AREA * 100) * (1 - current_hcc / K_VALS)
    growth  <- (pmax(0, R_GROW * (1 - p_vec))) * current_hcc * (1 - current_hcc / K_VALS)
    
    current_hcc <- (current_hcc + growth + recruit) * (1 - EXP_MORT)
    current_hcc <- pmin(K_VALS, pmax(0, current_hcc))
    ts_matrix[, t] <- current_hcc
  }
  return(ts_matrix)
}
```

### RUN INDIVIDUALIZED SCENARIOS (2-4: Top Beneficiaries Discovery)

```{r}
cat("Calculating S1 Baseline...\n")
s1_ts <- run_core_logic_ts(HCC_INIT, STEPS, PENALTY)

ph_indices <- which(IS_PH)
n_cores <- max(1, detectCores() - 1)
cl <- makeCluster(n_cores)
registerDoParallel(cl)

clusterExport(cl, c("run_core_logic_ts", "P_T", "HCC_INIT", "PENALTY", "AREA", "K_VALS",
                    "R_GROW", "EXP_MORT", "LARVAE_PER_HCC", "MATURITY_RATE",
                    "COLONY_TO_KMSQ", "STEPS", "RESTORATION_LARVAE"))
clusterEvalQ(cl, library(Matrix))

cat(paste("Running Localized Hotspot Analysis (S2-S4) on", n_cores, "cores...\n"))

individual_results <- foreach(idx = ph_indices, .options.snow = list(preschedule = TRUE)) %dopar% {
  local_penalty <- PENALTY
  local_penalty[idx] <- 0
  
  s2_ts <- run_core_logic_ts(HCC_INIT, STEPS, PENALTY, r_indices = idx, r_larvae = RESTORATION_LARVAE)
  s3_ts <- run_core_logic_ts(HCC_INIT, STEPS, local_penalty)
  s4_ts <- run_core_logic_ts(HCC_INIT, STEPS, local_penalty, r_indices = idx, r_larvae = RESTORATION_LARVAE)
  
  list(
    final = c(s2_ts[idx, STEPS], s3_ts[idx, STEPS], s4_ts[idx, STEPS]),
    ts    = rbind(s2_ts[idx, ], s3_ts[idx, ], s4_ts[idx, ])
  )
}
stopCluster(cl)
```

### RUN PORTFOLIO SCENARIOS (5-7: Group Intervention)

```{r}

cat("Calculating Portfolio Scenarios (5-7)...\n")
portfolio_penalty <- PENALTY
portfolio_penalty[target_indices] <- 0

s5_ts <- run_core_logic_ts(HCC_INIT, STEPS, PENALTY, r_indices = target_indices, r_larvae = RESTORATION_LARVAE)
s6_ts <- run_core_logic_ts(HCC_INIT, STEPS, portfolio_penalty)
s7_ts <- run_core_logic_ts(HCC_INIT, STEPS, portfolio_penalty, r_indices = target_indices, r_larvae = RESTORATION_LARVAE)
```

## Build Output Dataframes

```{r}
ph_reefs <- REEF[IS_PH]
n_ph     <- sum(IS_PH)

# Individual gain data at end of simulation (S2-S4)
indiv_reefgain_df <- data.frame(
  reefid = REEF[ph_indices],
  S1_baseline = s1_ts[ph_indices, STEPS],
  do.call(rbind, lapply(individual_results, `[[`, "final")) %>% `colnames<-`(c("S2_active", "S3_passive", "S4_combined"))
) %>% mutate(
  active_gain = S2_active - S1_baseline,
  passive_gain = S3_passive - S1_baseline,
  combined_gain = S4_combined - S1_baseline
)

# Individual HCC time series (S2-S4): HCC of each reef when it was the sole target
indiv_ts_df <- bind_rows(lapply(seq_along(ph_indices), function(i) {
  ts <- individual_results[[i]]$ts
  data.frame(
    reefid      = REEF[ph_indices[i]],
    year        = 1:STEPS,
    S2_active   = ts[1, ],
    S3_passive  = ts[2, ],
    S4_combined = ts[3, ]
  )
}))

# Archives all yearly HC for all PH reefs. Note that S2-S4 indicates HC if individual intervention while S5-S7 are for group interventions based on preselected_ids 
reef_yearly_ts <- data.frame(
  reefid      = rep(ph_reefs, times = STEPS),
  year        = rep(1:STEPS, each = n_ph),
  S1_baseline = as.vector(s1_ts[IS_PH, ]),
  S2_active   = NA_real_,
  S3_passive  = NA_real_,
  S4_combined = NA_real_,
  S5_active   = as.vector(s5_ts[IS_PH, ]),
  S6_passive  = as.vector(s6_ts[IS_PH, ]),
  S7_combined = as.vector(s7_ts[IS_PH, ]),
  is_targeted = rep(ph_reefs %in% preselected_ids, times = STEPS)
) %>% rows_update(indiv_ts_df, by = c("reefid", "year"), unmatched = "ignore")


# End of simulation gain for S5-S7
group_reefgain_df <- data.frame(
  reefid  = ph_reefs,
  S5_gain = s5_ts[IS_PH, STEPS] - s1_ts[IS_PH, STEPS],
  S6_gain = s6_ts[IS_PH, STEPS] - s1_ts[IS_PH, STEPS],
  S7_gain = s7_ts[IS_PH, STEPS] - s1_ts[IS_PH, STEPS]
)
```

## Visualization

### Transformation to long form

```{r}
# indiv_reefgain_df transform for visualization
indiv_reefgain_long <- indiv_reefgain_df %>%
  select(reefid, active_gain, passive_gain, combined_gain) %>%
  pivot_longer(-reefid, names_to = "intervention", values_to = "gain") %>%
  mutate(intervention = recode(intervention,
                               active_gain   = "S2 Active",
                               passive_gain  = "S3 Passive",
                               combined_gain = "S4 Combined"
  ))

# group_reefgain_df transform for visualization
group_reefgain_long <- group_reefgain_df %>%
  pivot_longer(-reefid, names_to = "intervention", values_to = "gain") %>%
  mutate(intervention = recode(intervention,
                               S5_gain = "S5 Active",
                               S6_gain = "S6 Passive",
                               S7_gain = "S7 Combined"
  ))

# group_reefgain_df transform for visualization (targeted reefs only)
group_reefgain_targeted <- group_reefgain_df %>%
  filter(reefid %in% preselected_ids) %>%
  pivot_longer(-reefid, names_to = "intervention", values_to = "gain") %>%
  mutate(intervention = recode(intervention,
                               S5_gain = "S5 Active",
                               S6_gain = "S6 Passive",
                               S7_gain = "S7 Combined"
  ))
```

### Boxplots

```{r}

# Individual reef HCC gain vs baseline — all PH reefs (S2-S4)
indiv_reefgain_ph_boxplot <- ggplot(indiv_reefgain_long, aes(x = intervention, y = gain, fill = intervention)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  labs(title = "Individual Reef Potential (S2-S4)", y = paste0("Gain in HCC (Yr ", STEPS, ")")) +
  theme_minimal()

# Network-wide HCC gain vs baseline — all PH reefs (S5-S7)
group_reefgain_ph_boxplot <- ggplot(group_reefgain_long, aes(x = intervention, y = gain, fill = intervention)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  labs(title = "Portfolio Network-Wide Gain vs Baseline (S5-S7)",
       y = paste0("HCC Gain (Yr ", STEPS, ")")) +
  theme_minimal()

# HCC gain vs baseline — targeted portfolio reefs only (S5-S7)
group_reefgain_target_boxplot <- ggplot(group_reefgain_targeted, aes(x = intervention, y = gain, fill = intervention)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  labs(title = paste0("Portfolio Gain — Targeted Reefs (S5-S7 vs S1, n=", length(preselected_ids), ")"),
       y = paste0("HCC Gain (Yr ", STEPS, ")")) +
  theme_minimal()

print(indiv_reefgain_ph_boxplot)
print(group_reefgain_ph_boxplot)
print(group_reefgain_target_boxplot)
```

## Export

Please make sure that there is a folder in the project folder named "output".

```{r}
output_file <- paste0("output/df_ph_yearly_hc-", format(Sys.Date(), "%Y%m%d"), ".xlsx")
write.xlsx(
  list(
    "Yearly_HCC"                = reef_yearly_ts,
    "Individual Intervention Gain" = indiv_reefgain_df,
    "Group Intervention Gain"   = group_reefgain_df
  ),
  file = output_file
)
cat(paste("Exported:", output_file, "\n"))
cat("Analysis DONE!\n")
```
